---
output: github_document
---

# Economic Self-Reliance and Gender Inequality

This is a replication repository for Deirdre Bloome, Derek Burk, and Leslie 
McCall, "Economic Self-Reliance and Gender Inequality between U.S. Men and 
Women, 1970–2010," American Journal of Sociology 124, no. 5 (March 2019): 1413-
1467.

## Overview of directory structure

**main**: This directory contains all the scripts that can be used to reproduce 
our main analysis.

**functions**: This directory contains scripts which define functions used in 
the analysis scripts in **main** and **auxiliary**.

**original_data**: This directory contains all data files necessary to 
reproduce our analyses. All data are derived from IPUMS CPS extracts, downloaded 
from the [IPUMS CPS website](https://cps.ipums.org/cps/). These data are posted 
here with the permission of IPUMS.

Sarah Flood, Miriam King, Steven Ruggles, and J. Robert Warren. Integrated
Public Use Microdata Series, Current Population Survey: Version 5.0 [dataset].
Minneapolis, MN: University of Minnesota, 2017. https://doi.org/10.18128/D030.V5.0

The file "cps_master.Rdata" was created by loading an IPUMS extract downloaded 
as a .dta file into R using the **foreign** package, and saving it to an .Rdata 
file without modification. All other data files were downloaded directly from 
IPUMS CPS. The file "topcode_values.csv" is derived from the income component 
topcode values provided [here](https://cps.ipums.org/cps/topcodes_tables.shtml).

**auxiliary**: This directory contains scripts used for sensitivity analyses.

## Required R packages

The following packages were used in our analysis:

* data.table
* stringr
* foreign
* mice, version 2.25
* nnet
* car
* MASS
* pscl
* RCurl
* dplyr
* magrittr
* ipumsr
* Hmisc
* weights

Unfortunately, we did not rigorously track the package versions used in the 
analysis, with the exception of the mice package. However, we were able to 
replicate our results with the following setup:

```{r warning=FALSE}
pkgs <- c("data.table", "stringr", "foreign", "mice", "nnet", "car", "MASS", 
          "pscl", "RCurl", "dplyr", "magrittr", "ipumsr", "Hmisc", "weights")
for (p in pkgs) {
    suppressPackageStartupMessages(library(p, character.only = TRUE))
}
sessionInfo()
```

To install mice version 2.25, you can use the following code:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_version("mice", version = "2.25")
```



## Replication guide

All scripts necessary to reproduce our main analysis can be found in the 
**main** directory. The subdirectories of **main** are numbered according to the
order in which scripts should be run, and the files within each subdirectory 
are similarly numbered. We list these scripts below, in order, and include 
notes describing any parameters that need to be set and the outputs created by 
each script. All scripts create output in the same directory as the script 
itself, or in a subdirectory of that directory.

### main/1_clean_data

* **clean_data_step_1.R**: Creates output "cleaned_data_step_1.Rdata".
* **clean_data_step_2.R**: Creates output "cleaned_data_step_2.Rdata".
* **clean_data_step_2_no_imputation**: Creates output "cleaned_data_step_2_no_imputation.Rdata".
* **clean_data_step_3.R**: This script requires the analyst to set the value of 
  the `imputation` parameter at the top of the script to `TRUE` or `FALSE`. 
  When `imputation == TRUE`, the script creates output "cleaned_data_step_3.Rdata". 
  When `imputation == FALSE`, the script creates output "cleaned_data_step_3_no_imputation.Rdata".
* **clean_data_step_4.R**: This script requires the analyst to set the value of 
  the `imputation` parameter at the top of the script to `TRUE` or `FALSE`. 
  When `imputation == TRUE`, the script creates output "cleaned_data_step_4.Rdata". 
  When `imputation == FALSE`, the script creates output "cleaned_data_step_4_no_imputation.Rdata".
* **clean_data_step_5.R**: This script requires the analyst to set the value of 
  the `imputation` parameter at the top of the script to `TRUE` or `FALSE`. 
  When `imputation == TRUE`, the script creates output "cleaned_data_step_5.Rdata". 
  When `imputation == FALSE`, the script creates output "cleaned_data_step_5_no_imputation.Rdata".
  
### main/2_prepare_for_imputation

In selecting variables to include in the imputation, we followed recommendations
from:

van Buuren, Stef Flexible Imputation of Missing Data. CRC Press, 2012.
(particularly Chapter 5)

van Buuren, S., and Karin Groothuis-Oudshoorn. "mice: Multivariate imputation by 
chained equations in R." Journal of statistical software (2010): 1-68.

* **1_examine_distributions_pre_imputation.R**: This script creates histograms
  of all variables to assess the need to transform variables included in the 
  multiple imputation. The histograms are saved as .png files in the 
  "1_variable_distributions" subdirectory. We visually examined the histograms to 
  decide which variables to transform, and recorded our decisions in the 
  file "1_variable_distributions/variable_transformations.csv", which is called 
  upon in the next script.
* **2_get_corr.sh**: This script cannot be run as-is, but is included for 
  documentary purposes. It was used to submit the script 
  "2_get_corr_r2_cramers_v_by_decade_array.R" to a cluster job runner in order 
  to run it separately on each decade of our data.
* **2_get_corr_r2_cramers_v_by_decade_array.R**: This script was written to be 
  run as part of a cluster job that would operate on each decade of data in 
  parallel, but can be run locally by setting the `NAI` parameter at the top 
  of the file to a value between 1 and 5, corresponding to the five decades 
  between 1970 and 2010, or by supplying the value as a trailing argument 
  if running the script from the command line. Also note that this script may 
  print error messages, but as long as execution does not stop, these errors 
  have been handled (with the `try` function) and can be safely ignored.
* **3_create_pred_matrix_by_decade.R**: Creates output 
  "3_pred_matrix_list.Rdata".
  
### main/3_multiply_impute

* **1_full_imputation_by_year_array_with_ppc_and_inf_check.R**: This script was
  written to be run as part of a cluster job that would operate on each decade
  of data in parallel, but can be run locally by setting the `NAI` parameter at
  the top of the file to a value between 1 and 5, corresponding to the five
  decades between 1970 and 2010, or by supplying the value as a trailing
  argument if running the script from the command line. For each decade, this 
  script creates 10 imputed datasets, and saves a file containing each of 10 
  iterations along the way so that not all progress is lost if the script is 
  interrupted, in the subdirectory "1_imp_iterations". If the script is 
  interrupted, you can simply run it again and it will pick up where it left 
  off. The final imputation results will all be contained in 
  "imp_YEAR_10.Rdata", the file for the 10th iteration.
  The script also creates diagnostic output in the "output" subdirectory, and 
  in the case of an execution error, saves degugging information in the 
  "1_error_dump" subdirectory.
  This is a memory- and computationally-intensive step in the 
  analysis. Creating the 10 imputed datasets for 1970, which has the fewest 
  observations and the least amount of missingness, required about 6 GB of RAM 
  and took around 25 hours. Creating the imputed datasets for 2010 required
  about 14 GB of RAM and took over two weeks. 
  This long run time is mostly due to our use of a 
  two-stage imputation process for variables with a high rate of zero values 
  (which includes many of our income variables), where the first stage is a 
  computationally-intensive logistic regression.
* **2_assess_extreme_imputed_values.R**: Creates five files with filenames of 
  the form "2_extreme_values_YEAR.csv", which contain information on extreme 
  values imputed for topcoded values of income component variables. This output
  was used to help set an upper threshold for imputed values, which is 
  implemented in the next script.
* **3_enforce_upper_thresholds_on_imputed_values.R**: This script compresses 
  the distribution of imputed, topcoded income component values for variables 
  with extreme imputed values so that all such values fall below a defined 
  threshold. It produces five files with filenames of the form 
  "3_imp_YEAR_10_extreme_values_transposed.Rdata", as well as diagnostic files
  of the form "3_extreme_values_transposition_report_YEAR.csv".
  
### main/4a_estimate_taxes_and_transfers_imputed

* **1_prepare_taxsim_input_files.R**: Creates subdirectory "1_taxsim_input" and 
  files in that subdirectory for each imputed dataset for each decade, with 
  filenames of the form "srYEAR_IMPNUM" (e.g., "sr1970_1").
* **2_check_dependent_counts.R**: This script checks for a particular problem 
  we encountered on early TAXSIM runs to make sure that our fix worked. It 
  produces no output. Instead, it prints a message to the console indicating 
  whether the problem is fixed.
* **3_upload_taxsim_input.R**: This script uploads the files created in step 1 
  to the TAXSIM ftp server at taxsimftp.nber.org. If the script isn't working 
  for unclear reasons, make sure your firewall settings allow ftp see 
  <http://users.nber.org/~taxsim/ftp-problems.html>.
  Also, for more info on the TAXSIM ftp service, see 
  <https://users.nber.org/~taxsim/taxsim9/taxsim-ftp.html>.
* **4_download_taxsim_output.R**: This script downloads the TAXSIM output, 
  saving the output files in the "4_taxsim_output" directory. The TAXSIM 
  calculations are performed during file download, so this script can be run 
  immediately after your input files are uploaded. Note that the TAXSIM ftp 
  server deletes old files daily, so this script must be run within one day of 
  uploading a given file.
* **5_check_taxsim_input_output.r**: This script simply checks that all the 
  TAXSIM output files have the same number of lines as the corresponding input 
  files.
* **6_label_taxsim_output_and_merge_imputed.r**: This script creates five 
  output files, one for each decade, with filenames of the form 
  "6_imp_post_tax_YEAR.Rdata".
  
### main/4b_estimate_taxes_and_transfers_non_imputed

* **1_prepare_taxsim_input_files.R**: Creates subdirectory "1_taxsim_input" and 
  one output file ("sr1970_2010") containing all the TAXSIM input information.
* **2_check_dependent_counts.R**: This script checks for a particular problem 
  we encountered on early TAXSIM runs to make sure that our fix worked. It 
  produces no output. Instead, it prints a message to the console indicating 
  whether the problem is fixed.
* **3_upload_taxsim_input.R**: This script uploads the file created in step 1 
  to the TAXSIM ftp server at taxsimftp.nber.org. If the script isn't working 
  for unclear reasons, make sure your firewall settings allow ftp see 
  <http://users.nber.org/~taxsim/ftp-problems.html>.
  Also, for more info on the TAXSIM ftp service, see 
  <https://users.nber.org/~taxsim/taxsim9/taxsim-ftp.html>.
* **4_download_taxsim_output.R**: This script downloads the TAXSIM output, 
  saving the output file in the "4_taxsim_output" directory. The TAXSIM 
  calculations are performed during file download, so this script can be run 
  immediately after your input files are uploaded. Note that the TAXSIM ftp 
  server deletes old files daily, so this script must be run within one day of 
  uploading a given file.
* **5_check_taxsim_input_output.r**: This script simply checks that all the 
  TAXSIM output files have the same number of lines as the corresponding input 
  files.
* **6_label_taxsim_output_and_merge_non_imputed.r**: This script creates the 
  file "6_non_imp_data_post_tax.Rdata".
  
### main/5_make_no_cohab_datasets

* **1_make_no_cohab_datasets.R**: This script creates five output files, one 
  for each decade of imputed data, with filenames of the form 
  "1_imp_YEAR_post_tax_no_cohab.Rdata".
  
### main/6a_make_analysis_dataset_imputed

* **1_make_analysis_dataset_imputed.R**: This script creates five output files, 
  one for each decade of imputed data, with filenames of the form 
  "1_imps_1970_analysis_vars.Rdata".
  
### main/6b_make_analysis_dataset_non_imputed

* **1_make_analysis_dataset_non_imputed.R**: This script creates one output file
  named "1_non_imputed_analysis_vars.Rdata".
  
### main/6c_make_analysis_dataset_no_cohab

* **1_make_analysis_dataset_no_cohab.R**: This script creates five output files, 
  one for each decade of imputed data, with filenames of the form 
  "1_imps_1970_analysis_vars.Rdata".
  
### main/7_make_exclusion_flags

* **1_make_top_two_pct_flags.R**: This script creates two output files named  
  "1_top_2_pct_flag_non_imp.Rdata" and "1_top_2_pct_flag_imp.Rdata".

* **2_make_exclude_alloc_flag.R**: This script creates one output file named 
  "2_exclude_alloc_flag.Rdata".
  
### main/8a_perform_decomposition_imputed

* **1_make_decomp_component_tables.R**: This script allows you to adjust some 
  options by setting the following variables near the top of the file:
    * `fam_adj`: Should family income be adjusted for family size?
    * `exclude_alloc`: Should all cases where the focal person's or their 
      spouse's labor earnings were allocated or imputed be excluded from the 
      analysis?
    * `exclude_top_2_pct`: Should all families in the top 2% of family income 
      be excluded from the analysis?
    * `exclude_top_decile_female_earners`: Should all families containing a top 
      decile female earner be excluded?
    * `exclude_top_decile_male_earners`: Should all families containing a top 
      decile male earner be excluded?
  In our main analysis results, only `fam_adj` is set to `TRUE`. All other 
  options were only used for sensitivity analyses. This script produces ten 
  output files, one for each imputed dataset, in the subdirectory 
  "1_decomp_component_tables", with filenames of the form 
  "decomp_components_imputed_SUFFIX_IMPNUM.csv", where SUFFIX indicates the 
  options that were set to `TRUE`.
* **2_make_qois_for_tables_and_figs.R**: This script allows you to adjust the 
  same options described above, and produces five output files in the 
  "2_qois_for_tables_and_figs" subdirectory, with filenames of the form 
  "qoi_imp_YEAR_SUFFIX.Rdata".
* **3_make_figures_1_2_and_3.R**: This script allows you to adjust the same 
  options described above, and produces three output files in the 
  "3_figures_1_2_and_3" directory, with filenames 
  "figure_1_SUFFIX.csv", "figure_2_SUFFIX.csv", and "figure_3_SUFFIX.csv".

### main/8b_perform_decomposition_non_imputed

* **1_make_decomp_component_tables.R**: This script allows you to adjust the
  options described above, and produces one output file, in the subdirectory 
  "1_decomp_component_tables", with filename
  "decomp_components_non_imputed_SUFFIX.csv", where SUFFIX indicates the 
  options that were set to `TRUE`.
* **2_make_qois_for_tables_and_figs.R**: This script allows you to adjust the 
  same options described above, and produces five output files in the 
  "2_qois_for_tables_and_figs" subdirectory, with filenames of the form 
  "qoi_non_imp_YEAR_SUFFIX.Rdata".
* **3_make_figures_1_2_and_3.R**: This script allows you to adjust the same 
  options described above, and produces three output files in the 
  "3_figures_1_2_and_3" directory, with filenames 
  "figure_1_SUFFIX.csv", "figure_2_SUFFIX.csv", and "figure_3_SUFFIX.csv".

### main/8c_perform_decomposition_no_cohab

* **1_make_decomp_component_tables.R**: This script allows you to adjust the
  options described above, and produces ten 
  output files, one for each imputed dataset, in the subdirectory 
  "1_decomp_component_tables", with filenames of the form 
  "decomp_components_imputed_SUFFIX_IMPNUM.csv", where SUFFIX indicates the 
  options that were set to `TRUE`.
* **2_make_qois_for_tables_and_figs.R**: This script allows you to adjust the 
  same options described above, and produces five output files in the 
  "2_qois_for_tables_and_figs" subdirectory, with filenames of the form 
  "qoi_imp_YEAR_SUFFIX.Rdata".
* **3_make_figures_1_2_and_3.R**: This script allows you to adjust the same 
  options described above, and produces three output files in the 
  "3_figures_1_2_and_3" directory, with filenames 
  "figure_1_SUFFIX.csv", "figure_2_SUFFIX.csv", and "figure_3_SUFFIX.csv".
  
### main/9_make_tables

* **1_make_tables.R**: This script allows you to adjust the 
  same options described above, and produces one output file, 
  "1_tabs_list.Rdata", which contains the figures contained in all tables in the 
  paper.

* **2_bootstrap_loop_tables.R**: This script allows you to adjust the 
  same options described above, and produces two output files, 
  "2_all_bootstrap_decomp_tables.Rdata" and "2_tabs_list.Rdata". The file 
  "2_all_bootstrap_decomp_tables.Rdata" just contains intermediate output in 
  case the script hits an error or is interrupted. The file "2_tabs_list.Rdata" 
  contains the figures needed to compute bootstrapped uncertainty intervals for
  all tables in the paper